---
title: "PlotSvalbard - User Manual"
author: "Mikko Vihtakari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PlotSvalbard - User Manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[**PlotSvalbard**](https://github.com/MikkoVihtakari/PlotSvalbard) package provides functions to plot research data from Svalbard on high resolution maps in [R](https://www.r-project.org/). The package is based on [ggplot2](http://ggplot2.tidyverse.org/reference/) and the functions can be expanded using ggplot syntax. The package contains also maps from other places in the Arctic, including polar stereographic maps of the Arctic.

# Basemap

The `basemap` function 

## Map types

### Svalbard

```{r}
library(PlotSvalbard)
#basemap("svalbard")
```

#### Kongsfjorden

```{r}
basemap("kongsfjorden") # or just basemap()
```

```{r}
basemap("kongsfjordbotn")
```

```{r}
basemap("kronebreen")
```

```{r}
basemap("mosj")
```

### Barents Sea

```{r}
basemap("barentssea")
```

Barents Sea map also prints mainland Norway, but the projection is not optimal, and the resolution is quite low.

```{r}
basemap("barentssea", limits = c(12, 24, 68, 71))
```

### Pan-Arctic

To be added

## Adding data to basemaps

## Bathymetry

To be added

<!-- # Stuff to be deleted -->

<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output:  -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side.  -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->

<!-- You can enable figure captions by `fig_caption: yes` in YAML: -->

<!--     output: -->
<!--       rmarkdown::html_vignette: -->
<!--         fig_caption: yes -->

<!-- Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**. -->

<!-- ## More Examples -->

<!-- You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`. -->

<!-- ```{r, echo=FALSE, results='asis'} -->
<!-- knitr::kable(head(mtcars, 10)) -->
<!-- ``` -->

<!-- Also a quote using `>`: -->

<!-- > "He who gives up [code] safety for [code] speed deserves neither." -->
<!-- ([via](https://twitter.com/hadleywickham/status/504368538874703872)) -->
