---
title: "PlotSvalbard - User Manual"
subtitle: "Plot research data from Svalbard on maps "
author:
  name: Mikko Vihtakari
  email: mikko.vihtakari@gmail.com
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    toc: true
    theme: lumen
    toc_float: true
    toc_depth: 5
    collapsed: false
    variant: markdown+simple_table
    fig_caption: true
    number_sections: false
---

<!-- --- -->
<!-- title: "PlotSvalbard - User Manual" -->
<!-- author: "Mikko Vihtakari" -->
<!-- date: "`r Sys.Date()`" -->
<!-- output: rmarkdown::pdf_document -->
<!-- vignette: > -->
<!--   %\VignetteIndexEntry{PlotSvalbard - User Manual} -->
<!--   %\VignetteEngine{knitr::rmarkdown} -->
<!--   %\VignetteEncoding{UTF-8} -->
<!-- --- -->


```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.dpi = 96, collapse = TRUE, fig.width = 7, fig.height = 6, comment = "#>")
```

[**PlotSvalbard**](https://github.com/MikkoVihtakari/PlotSvalbard) package provides functions to plot research data from Svalbard on high resolution maps in [R](https://www.r-project.org/). The package is based on [ggplot2](http://ggplot2.tidyverse.org/reference/) and the functions can be expanded using ggplot syntax. The package contains also maps from other places in the Arctic, including polar stereographic maps of the Arctic.

<!-- Note that the quality of graphics in this document is restricted due to vignette requirements for R. The quality of maps printed on your graphics window should be much higher than here.  -->
The maps in PlotSvalbard are designed to be printed on `pdf()` device. If you need the maps for other devices, see the [Graphical parameters](#par) section.

# Basemap

The `basemap` function is the generic plotting command in PlotSvalbard, and is analogous to empty `ggplot()` call. The `basemap()` call plots a map that is specified by the `type` argument. Data that contains geographic information can be plotted on these maps using the [ggplot2 layers separated by the `+` operator](http://ggplot2.tidyverse.org/reference/).

## Map types

Map types are specified by the `type` argument in `basemap()` function. The list below shows all currently implemented map types. 

### Svalbard

```{r message=FALSE, warning=FALSE}
library(PlotSvalbard)
basemap("svalbard")
```

#### Kongsfjorden

```{r message=FALSE, warning=FALSE}
basemap("kongsfjorden") # or just basemap()
```

```{r message=FALSE, warning=FALSE}
basemap("kongsfjordbotn")
```

```{r message=FALSE, warning=FALSE}
basemap("kronebreen")
```

```{r message=FALSE, warning=FALSE}
basemap("mosj")
```

### Barents Sea

```{r message=FALSE, warning=FALSE}
basemap("barentssea")
```

Barents Sea map also prints mainland Norway, but the projection is not optimal, and the resolution is quite low.

```{r message=FALSE, warning=FALSE}
basemap("barentssea", limits = c(12, 24, 68, 71))
```

### Pan-Arctic

```{r message=FALSE, warning=FALSE}
basemap("arctic50")
```

```{r message=FALSE, warning=FALSE}
basemap("arctic60")
```

## Limiting (zooming) the maps

Any `basemap` can be limited (or zoomed in) using the `limits` argument. The `limits` argument has to be a numeric vector specifying the minimum longitude, maximum longitude, minimum latitude and maximum latitude of the limiting area in decimal degrees. The order and length (4) of the vector is rigid. 

Note that some map types are already limited, so if you are looking for a map of Kongsfjorden, using `basemap("kongsfjorden")` might be just what you need.

```{r}
basemap("svalbard", limits = c(11, 18, 77.9, 78.85))
```

```{r}
basemap("arctic50", limits = c(3*10^6, -3*10^6, 3*10^6, -3*10^6))
```

## Bathymetry

All basemaps support bathymetry, but the resolution of bathymetry shapefiles is currently too low to plot bathymetry inside fjords. Bathymetry can be plotted using the `bathymetry` argument.

```{r}
basemap("arctic60", bathymetry = TRUE)
```


```{r}
basemap("barentssea", bathymetry = TRUE)
```


```{r}
basemap("kongsfjorden", bathymetry = TRUE)
```

## Adding data to basemaps

Data can be added to basemaps using the [`+` operator and layers for ggplot2](http://ggplot2.tidyverse.org/reference/). Below you will find some examples on how to add your research data on basemaps.

### Adding station labels (text)

Text can be added to basemaps using the `geom_text()` function:

```{r message=FALSE, warning=FALSE}
data("npi_stations")

basemap("svalbard", limits = c(3,24,78.5,82), round.lat = 1, round.lon = 2, 
  border.col.land = "#a9750d", border.col.glacier = "grey95") + 
  geom_text(data = npi_stations, aes(x = lon.utm, y = lat.utm, 
    label = Station), color = "red", fontface = 2)
```

Text size can be mapped to variables using the standard ggplot2 syntax:

```{r message=FALSE, warning=FALSE}
data("kongsfjord_moorings")

basemap("kongsfjorden", limits = c(11.3, 12.69, 78.85, 79.1), round.lat = 0.05, round.lon = 0.5) + 
  geom_text(data = kongsfjord_moorings, aes(x = lon.utm, y = lat.utm, 
  label = Mooring.name, color = Name), fontface = 2, 
  size = 25.4/72.27*8) # font size = 8, see Graphical parameters

```

### Adding piecharts

The [`scatterpie`](https://cran.r-project.org/web/packages/scatterpie/index.html) package allows relatively easy plotting of piecharts on maps. Extensions for ggplot2 work together with PlotSvalbard.

```{r}
data(zooplankton)

x <- transform_coord(zooplankton, lon = "Longitude", lat = "Latitude", bind = TRUE)

species <- colnames(x)[!colnames(x) %in% c("lon.utm", "lat.utm", "ID",
  "Longitude", "Latitude", "Total")]

library(scatterpie)

basemap("barentssea", limits = c(4, 24, 79.5, 83.5), round.lon = 2, round.lat = 1) + 
  geom_scatterpie(aes(x = lon.utm, y = lat.utm, group = ID, r = 100*Total), 
  data = x, cols = species, size = 0.1) + scale_fill_discrete(name = "Species", 
  breaks = species, labels = parse(text = paste0("italic(" , sub("*\\.", "~", species), ")")))
```

### Adding interpolated 2D surfaces

PlotSvalbard uses the [`krige` function](https://www.rdocumentation.org/packages/gstat/versions/1.1-5/topics/krige) from the [**gstat** package](https://cran.r-project.org/web/packages/gstat/index.html) to produce interpolated 2D plots:

```{r}
data("chlorophyll")

x <- interpolate(chlorophyll, Subset = "From <= 10", value = "Chla") ## Interpolate

plot(x, type = "kongsfjorden", limits = c(11.4,12.7,78.855,79.05), 
  legend.label = "Chlorophyll-a\n(mg/m3)") 

```

## Graphical parameters {#par}

The line widths and general looks of the basemaps are optimized for a half page sized `pdf` figure in scientific journals ([Springer figure dimensions](http://www.springer.com/authors/manuscript+guidelines?SGWID=0-40162-6-795162-0) were used to develop the function). The line widths may not look good when printed on other devices. You can modify the line widths and colors using `size.*` and `border.col.*` arguments.

```{r message=FALSE, warning=FALSE}
basemap("svalbard", size.land = 0.01, size.glacier = 0.05, size.grid = 0.05, 
  border.col.glacier = "#52bfe4", border.col.land = "#a9750d")
```

Approach to remove borders of land and glacier shapes:

```{r message=FALSE, warning=FALSE}
basemap("kongsfjorden", border.col.glacier = "grey95", border.col.land = "#eeeac4")
```

The [line width of ggplot2 is 2.13 wider than the line widths measured as points](https://stackoverflow.com/questions/47519624/how-is-the-line-width-size-defined-in-ggplot2/47521808#47521808) (pt). This means that if you want a certain line width, multiply the desired line width by $1/2.13$ inside `size.*` arguments. Similar conversion factor for font size is $1/2.845276$. 

## Performance

The `basemap("svalbard")` is currently fairly slow due to less than optimized code and the large size of the shapefiles that are used to generate the map. Zoomed in maps are considerably faster than full scale Svalbard maps.

If you are looking for optimal `limits` for your data, you can use the `basemap("barentssea")`
option to find these limits and replace `"barentssea"` with `"svalbard"` once you are done:

```{r}
system.time(basemap("barentssea"))
system.time(basemap("svalbard"))
system.time(basemap("barentssea", limits = c(c(19.5,23.5,80,81.7))))
system.time(basemap("svalbard", limits = c(c(19.5,23.5,80,81.7))))
```

# Citing PlotSvalbard

The `citation` function tells how to cite PlotSvalbard package

```{r message=FALSE, warning=FALSE}
citation("PlotSvalbard")
```

However, please note that the maps generated by this package should be cited to their original source. 

- Svalbard maps are property of the [Norwegian Polar Institute](http://geodata.npolar.no/)
- Barents Sea and pan-Arctic maps are downloaded from [Natural Earth Data](http://www.naturalearthdata.com/downloads/10m-physical-vectors/10m-land/)
- Bathymetry shapefiles are generalized from [IBCAO v3.0 500m RR grid](https://www.ngdc.noaa.gov/mgg/bathymetry/arctic/ibcaoversion3.html)

The example data included in the package are property of the Norwegian Polar Institute and should not be used in other instances. I.e. these data are unpublihed at the moment.

PlotSvalbard depends on many packages (see [**Dependencies**](https://mikkovihtakari.github.io/PlotSvalbard/index.html#dependencies)). If you use these packages in your publications, please cite the respective packages.